# Datasets in R {#dataset} 

There are various pre-built data sets included with R, which are typically used as practice sets for playing with R functions. We'll first go through how to load and use built-in R data sets in this lecture. The most popular R demo data sets will then be discussed, including `mtcars`, `iris`, `ToothGrowth`, `PlantGrowth`, and `USArrests`.   

## Pre-loaded data
To see the list of pre-loaded data in your system, type the function `data()`:  

```{r, eval=FALSE}  
data()
```  
  
## Loading a built-in R data  

Load and print `mtcars` data as follow:

```{r, eval=TRUE}  
# Loading
data(mtcars)
# Print the first 6 rows
head(mtcars, 6)
```  



## Most used data sets in R  

### mtcars  
`mtcars` data was extracted from the 1974 Motor Trend US magazine, and comprises fuel consumption and 10 aspects of automobile design and performance for 32 automobiles (1973–74 models).  
  
```{r, eval=TRUE}  
# 1. Loading 
data("mtcars")  

# 2. Print
head(mtcars)  

# Number of rows (observations)
nrow(mtcars)  

# Number of columns (variables)
ncol(mtcars)
```  
If you want learn more about `mtcars` data sets, type this: `?mtcars`  
```{r, eval=FALSE}  
?mtcars
```   

### iris  

`iris` data set gives the measurements in centimeters of the variables sepal length, sepal width, petal length and petal width, respectively, for 50 flowers from each of 3 species of iris. The species are *Iris setosa*, *versicolor*, and *virginica*.  

```{r, eval=TRUE}  
data("iris")
head(iris)
```   
If you want learn more about `iris` data sets, type this: `?iris`  
```{r, eval=FALSE}  
?iris
```   


### ToothGrowth  

The results of an experiment examining the impact of vitamin C on tooth development in 60 Guinea pigs are included in the data set. Each rat received one of three vitamin C dose levels (0.5, 1, and 2 mg/day) using either ascorbic acid or orange juice as the delivery mechanism (a form of vitamin C and coded as VC).  

```{r, eval=TRUE} 
data("ToothGrowth")
head(ToothGrowth)
```  

If you want learn more about `ToothGrowth` data sets, type this: `?ToothGrowth`  
```{r, eval=FALSE}  
?ToothGrowth
```   


### PlantGrowth
Results from an experiment comparing yields between two distinct treatment conditions and a control condition (as determined by the dry weight of plants).

```{r, eval=TRUE} 
data("PlantGrowth")
head(PlantGrowth)
```  

### USArrests  

Statistics on the rates of violent crime in each US state  

```{r, eval=TRUE} 
data("USArrests")
head(USArrests)
```  








