# R basics  

In this section we will be dealing with basic operations using R. One should be aware of these basics before proceeding to any kind of data analysis using R.  

The RStudio interface is simple. You type R code into the bottom line of the RStudio console pane (see figure \@ref(fig:rstudio2)) and then click Enter to run it. The code you type is called a command, because it will command your computer to do something for you. The line you type it into is called the command line.  

>'#' hashtag in R  
R treats the hashtag character, '#'in a special way; R will not run anything that follows a hashtag on a line. This makes hashtags very useful for adding comments and annotations to your code. We will be able to read the comments, but your computer will not process it.

## Basic arithmetic operations  

```{r}
# Try these codes in R, Type these codes in console and hit enter
# result of R code is also shown here  

# + (Addition)   
7 + 4  

# - (Subtraction)  
7 - 4  

# * (Multiplication)
7 * 2  

# / (division) 
7 / 2  

# ^ (exponentiation) 
7 ^ 2
```  

The colon operator (`:`) returns every integer between two integers. It is an easy way to create a sequence of numbers.  

```{r}
100:130
```  


## Basic arithmetic functions:  







 

