# R basics  

In this section we will be dealing with basic operations using R. One should be aware of these basics before proceeding to any kind of data analysis using R.  

The RStudio interface is simple. You type R code into the bottom line of the RStudio console pane (see figure \@ref(fig:rstudio2)) and then click Enter to run it. The code you type is called a command, because it will command your computer to do something for you. The line you type it into is called the command line.  

>'#' hashtag in R  
R treats the hashtag character, '#'in a special way; R will not run anything that follows a hashtag on a line. This makes hashtags very useful for adding comments and annotations to your code. We will be able to read the comments, but your computer will not process it.

## Basic arithmetic operations  

```{r}
# Try these codes in R, Type these codes in console and hit enter
# result of R code is also shown here  

# + (Addition)   
7 + 4  

# - (Subtraction)  
7 - 4  

# * (Multiplication)
7 * 2  

# / (division) 
7 / 2  

# ^ (exponentiation) 
7 ^ 2
```  

The colon operator (`:`) returns every integer between two integers. It is an easy way to create a sequence of numbers.  

```{r}
100:130
```  


## Basic arithmetic functions:  

```{r}  
# Logarithm to the base e  
log(4)  
# Logarithm to the base 10  
log10(4)  
# Logarithm to the base 2  
log2(4)  
# absolute value
abs(-4)  
# square root
sqrt(4) 

```


## Assigning values to variables  

```{r}  
x <- 2
# it is that value 2 is assigned to variable x  

y <- 5  

# it is that value 5 is assigned to variable y

# Now if you give x+5 or using any operator between x and y, you can see as below  

x + y  
x * y  
x / y  
x + 2*y  

```

> Note that R is case sensitive, i.e. x and X are not equal.  

## Basic data types  

+ numeric  
+ character  
+ logical  

```{r}
my_age <- 32   
# Numeric variable 32 is assigned to my_age

# while writing your code it is preffered to use '_' instead of space  

my_name <- "Dr Pratheesh" # Character variable  

#  Are you a data scientist?: (yes/no) <=> (TRUE/FALSE)  

is_datascientist <- TRUE # logical variable

```  

### Vectors  

A combination of multiple values (numeric, character or logical)  

#### How to create a vector {-}
Create a vector: c() for concatenate  

Case of missing values: **NA** (not available) and **NaN** (not a number)  

Get a subset of a vector: my_vector[i] to get the i^th^ element

```{r}  
# Create a numeric vector
student_ages <- c(27, 25, 29, 26, 20, 21, 23, 25)  

# Create a character vector
student_name <- c("asha", "adhi", "aravind", 
                  "mary", "peter", "daisy", 
                  "papu", "ramu")  

# subset of a vector  

# obtain 3 rd element from student_name  

student_name[3]

```


> Case of missing values: **NA** (not available) and **NaN** (not a number)
> Get a subset of a vector: my_vector[i] to get the ith element  

#### Calculations with vectors {-}  
> max(x), min(x), range(x), length(x), sum(x), mean(x), prod(x): product of the elements in x, sd(x): standard deviation, var(x): variance, sort(x)  

```{r}  
# Create a numeric vector
student_ages <- c(27, 25, 29, 26, 20, 21, 23, 25)  

# Maximum value of the vector
max(student_ages)

# Minimum value of the vector
min(student_ages)

# Range of the vector
range(student_ages)  

# Length of the vector
length(student_ages)  

# Total of the values in the vector
sum(student_ages)

# Mean of the vector
mean(student_ages)  

# Product of the elements in the vector
prod(student_ages)

# Standard deviation  
sd(student_ages)

# Variance of the vector
var(student_ages)

# Sort the values of the vector

# Ascending order
sort(student_ages,decreasing = FALSE)

# Descending order
sort(student_ages,decreasing = TRUE)
```
  
### Matrices  

It’s a homogeneous collection of data sets which is arranged in a two dimensional organisation. It’s a m*n array with similar data type. It is created using a vector input. It has a fixed number of rows and columns. You can perform many arithmetic operations on R matrix like – addition, subtraction, multiplication, and divisions.  

Create and naming matrix: matrix(), cbind(), rbind(), rownames(), colnames()
Transpose a matrix: t()
Dimensions of a matrix: ncol(), nrow(), dim()
Get a subset of a matrix: my_data[row, col]
Calculations with numeric matrices: rowSums(), colSums(), rowMeans(), colMeans()

```{r}
# make three vectors c1, c2 and c3
c1<-c(3,4,5)
c2<-c(7,8,9)
c3<-c(11,12,13)

# Creating matrix by binding column wise  
A <- cbind(c1,c2,c3)
print(A)

# Creating matrix by binding row wise  
B <- rbind(c1,c2,c3)
print(B)  

# Adding two matrices
A+B

# Simple element by element multiplication up to matrices.

A*B

# Matrix multiplication  

A%*%B

# Using matrix() function
# Elements are arranged sequentially by row.
M <- matrix(c(3:14), nrow = 4, byrow = TRUE)
print(M)

# Elements are arranged sequentially by column.
N <- matrix(c(3:14), nrow = 4, byrow = FALSE)
print(N)

# Define the column and row names.
rownames = c("row1", "row2", "row3", "row4")
colnames = c("col1", "col2", "col3")

P <- matrix(c(3:14), nrow = 4, byrow = TRUE, dimnames = list(rownames, colnames))
print(P)  

# Access the element at 3rd column and 1st row.
P[1,3]

# Access the element at 2nd column and 4th row.
P[4,2]

# Access only the  2nd row.
P[2,]

# Access only the 3rd column.  
P[,3]  


# Assign element at 1st row and 3rd column of matrix P to the variable x  

x<-P[1,3]  
print(x)

# create a vector from the second column of matrix P  

y<-P[,2]
y<-as.vector(y)

#Transpose a matrix
Q<-t(P)
print(Q)

# Knowing dimensions of a matrix

#Number of columns in A
ncol(A)  

#Number of rows in A
nrow(A)

# Number of rows and columns
dim(A)  

# Calculations with numeric matrices  

# Compute row sums of matrix A
rowSums(A)  

# Compute column sums of matrix A
colSums(A)  

# Compute row means of matrix A
rowMeans(A)  

# Compute column means of matrix A
colMeans(A)

```  


```{block2 item2, type='rmdnote'}
**Practical 1**: Matrices and vectors  

1. Create a vector of numbers form 23 to 33
2. Assign the vector with a name "numbers"  
3. Create a Matrix named "mat_num" with elements from 1 to 20 in 4 rows and 5 columns  
4. Name the columns and rows of matrix mat_num with any names you like  
5. Compute column means and row means of matrix mat_num
6. Access the element in second row and third column of matrix mat_num
``` 






