# Descriptive Statistics  

Before proceeding further please go through the first 6 chapters from my book [Textbook of Agricultural Statistics](https://kaumelon.github.io/stat3202/).  

You can download iris data from [here](https://drive.google.com/file/d/1AcAeL0XKOcYA3z99qmaQFvTr7gw3gxRB/view?usp=sharing) or use the inbuilt data set as discussed in chapter \@ref(dataset).  


To do any analysis we first import data, see Chapter \@ref(import). You can inspect your data using the functions `head()` and `tail()`, which will display the first and the last part of the data, respectively.

```{r, eval=TRUE}

# import data set in to R
# here we are using iris dataset

my_data<-iris # iris dataset is called in to variable my_data

# display first 6 rows
head(my_data, 6)  

# display last 6 rows
tail(my_data,6)

```  
## List of R functions  

**Descriptive Statistics**|**R Function**
:-----:|:-----:
Mean|mean()
Standard deviation|sd()
Variance|var()
Minimum|min()
Maximum|maximum()
Median|median()
Range of values (minimum and maximum)|range()
Sample quantiles|quantile()
Generic function|summary()
Interquartile range|IQR()  
  
The function `mfv()`, for most frequent value, [in `modeest` package] can be used to find the statistical mode of a numeric vector.  

## Measures of central tendancy  

```{r, eval=TRUE}
# Compute the mean value
mean(iris$Sepal.Length)

# Compute the median value
median(iris$Sepal.Length)  

# Compute the mode
# install.packages("modeest")
require(modeest)
mfv(iris$Sepal.Length)  

# Compute the minimum value
min(iris$Sepal.Length)

# Compute the maximum value
max(iris$Sepal.Length)  

# Quartiles  
quantile(iris$Sepal.Length)  

# Compute the median
median(iris$Sepal.Length)

```   

##  Measures of dispersion 

```{r, eval=TRUE}

# Range
# minimum value and maximum value is displayed
range(iris$Sepal.Length)  

# Inter quartile range  
IQR(iris$Sepal.Length)  

# Compute the variance
var(iris$Sepal.Length)

# Compute the standard deviation 
sd(iris$Sepal.Length)


# Compute the median absolute deviation
mad(iris$Sepal.Length)  

```  

## Overall summary   

The overall summary statistics of One variable or an entire data frame can be displayed using the function `summary()`.  


### Summary of single variable  

The mean, median, 25th and 75th quartiles, minimum and maximum values are all returned in a single line call.  

```{r, eval=TRUE}
summary(iris$Sepal.Length)
```

### Summary of a Data frame  

In this instance, each column receives an automated use of the summary() method. The kind of data in the column determines how the output is formatted. For instance:  

- The mean, median, min, max, and quartiles are returned if the column is a numeric variable.  

- The total number of observations in each group is returned if the column is a factor variable.

```{r, eval=TRUE}
summary(iris)
```   



## The sapply function  

The `sapply()` method may also be used to apply a specific function on a list or vector. For example, we may use it to compute the mean, sd, var, min, quantile,... for each column in a data frame.  


```{r, eval=TRUE}
sapply(iris[, -5], mean)
# The command direct to find the mean of all columns in the dataset iris except 5th column, which is species and is non numeric 

# Compute quartiles
sapply(iris[, -5], quantile)

```  

## Some other useful functions  

### stat.desc()

The function `stat.desc()` [in pastecs package], provides lot of useful statistics in a single call

```{r, eval=TRUE}
#Install 'pastecs' package
#load the library
library(pastecs)  

res <- stat.desc(my_data[, -5]) # results were stored to the variable 'res'  

# decimal digits were rounded to 2 using round()function
round(res, 2)  

```

## Missing values  

It should be noted that some R functions will return errors or NA even if just one item is missing or when the data contains missing values. For instance, if a vector has even a single missing value, the `mean()` method will return **NA**. The option `na.rm = TRUE`, which instructs the function to delete any NAs before computations, can be used to prevent this. Here's an example of how to use the mean function:

```{r, eval=FALSE}
mean(iris$Sepal.Length, na.rm = TRUE)
```  

## Frequency table  

Dataset below shows the number of children in 50 household collected as a part of survey. I will demonstrate how frequency table can be constructed.

```{r setup, eval=TRUE}
# Create example vector  
set.seed(99)
survey_data<- floor(runif(50, min=0, max=10)) 
survey_data                                                  
```
The frequency counts of the variables that appear in the chosen column of the dataframe are calculated using the `table()` function in R. The outcome is presented as a two-row tabular structure, with the first row displaying the value of each column and the second row displaying the matching frequencies.  

```{r, eval=TRUE}
# Frequency Table

freq_table<- table(survey_data)
freq_table

```  

### Cumulative frequency  
The total of all classes in a frequency distribution table, including this class below it, makes up the cumulative frequency distribution of a particular data set. The value at every cell location is calculated by adding the current value and all previous values encountered up to that point. This may be determined by using the `cumsum()` method.  

```{r, eval=TRUE}
# cumulative frequency

cumsum(freq_table)

``` 
From the above results now you can easily answer the question: number of families having 8 or less than 8 children?. It is 45  


## Grouped Frequency distribution  

```{r, eval=TRUE}

# Constucting frequency table
data<-iris$Sepal.Length  

# Decide how many classes you wish to use using struges formula
N<- length(data)
k<-1+3.322*log(N,base=10)
n_class<-ceiling(k) #Rounds value of k to nearest highest integer.
  

#Determine the class width  
c<-(max(data)-min(data))/k

# Rounded value of c 
class_width<-round(c,1) 

# Find lower limit  
L<-min(data)-((c*k-(max(data)-min(data)))/2)

              
# Upper limit  

U<-L+(class_width*n_class)




#to the closest half-integers, we come up with the interval [4, 8].
#Class width is set as 0.5
breaks = seq(L, U, by=class_width) 


data.cut = cut(data, breaks, right=FALSE)
data.freq = table(data.cut)
data.freq

```


```{block2 item6, type='rmdnote'}
**Self assesment**

1. Download the dataset by clicking [Here](https://drive.google.com/file/d/1_2U-_JbJlJyGOUhaDrTtXTGOgZZlx6Wx/view?usp=sharing)  
2. Find the descriptive statistics for the data using the functions above.  

```