# R basics  

In this section we will be dealing with basic operations using R. One should be aware of these basics before proceeding to any kind of data analysis using R.  

The RStudio interface is simple. You type R code into the bottom line of the RStudio console pane (see figure \@ref(fig:rstudio2)) and then click Enter to run it. The code you type is called a command, because it will command your computer to do something for you. The line you type it into is called the command line.  

>'#' hashtag in R  
R treats the hashtag character, '#'in a special way; R will not run anything that follows a hashtag on a line. This makes hashtags very useful for adding comments and annotations to your code. We will be able to read the comments, but your computer will not process it.

## Basic arithmetic operations  

```{r}
# Try these codes in R, Type these codes in console and hit enter
# result of R code is also shown here  

# + (Addition)   
7 + 4  

# - (Subtraction)  
7 - 4  

# * (Multiplication)
7 * 2  

# / (division) 
7 / 2  

# ^ (exponentiation) 
7 ^ 2
```  

The colon operator (`:`) returns every integer between two integers. It is an easy way to create a sequence of numbers.  

```{r}
100:130
```  


## Basic arithmetic functions:  

```{r}  
# Logarithm to the base e  
log(4)  
# Logarithm to the base 10  
log10(4)  
# Logarithm to the base 2  
log2(4)  
# absolute value
abs(-4)  
# square root
sqrt(4) 

```


## Assigning values to variables  

```{r}  
x <- 2
# it is that value 2 is assigned to variable x  

y <- 5  

# it is that value 5 is assigned to variable y

# Now if you give x+5 or using any operator between x and y, you can see as below  

x + y  
x * y  
x / y  
x + 2*y  

```

> Note that R is case sensitive, i.e. x and X are not equal.  

## Basic data types  

+ numeric  
+ character  
+ logical  

```{r}
my_age <- 32   
# Numeric variable 32 is assigned to my_age

# while writing your code it is preffered to use '_' instead of space  

my_name <- "Dr Pratheesh" # Character variable  

#  Are you a data scientist?: (yes/no) <=> (TRUE/FALSE)  

is_datascientist <- TRUE # logical variable

```  

### Vectors  

A combination of multiple values (numeric, character or logical)  

#### How to create a vector {-}
Create a vector: c() for concatenate  

Case of missing values: **NA** (not available) and **NaN** (not a number)  

Get a subset of a vector: my_vector[i] to get the i^th^ element

```{r}  
# Create a numeric vector
student_ages <- c(27, 25, 29, 26, 20, 21, 23, 25)  

# Create a character vector
student_name <- c("asha", "adhi", "aravind", 
                  "mary", "peter", "daisy", 
                  "papu", "ramu")  

# subset of a vector  

# obtain 3 rd element from student_name  

student_name[3]

```


> Case of missing values: **NA** (not available) and **NaN** (not a number)
> Get a subset of a vector: my_vector[i] to get the ith element  

#### Calculations with vectors {-}  
> max(x), min(x), range(x), length(x), sum(x), mean(x), prod(x): product of the elements in x, sd(x): standard deviation, var(x): variance, sort(x)

edit
